### 2013-09-03 Tex/chapters/introduction.tex
Rechtschreibfehler und Kleinigkeiten

The computing capabilities of modern computers enable researchers to
collect and analyze vast amounts of data. The prime example for this fact
is the LHC, where petabytes of particle collision data are collected,
processed and discarded or kept every second \cite{LHC}.
But computers are not only able to process measured data, but
also to generate data by simulations of e.g.\ statistical systems, where
the rules for each subsystem are well defined but the behavior of the
whole system is not easy to predict because of the interactions of the
subsystems. For those systems there often exists no analytic solutions
or only for very simplified or special cases. But one can simulate all
the interactions of the subsystems and observe the behavior of the whole
system.
For example to examine phase transitions, which are defined by the abrupt
change of an observable, e.g.\ the change of the density of water near
boiling or the change of the magnetization of a ferromagnet near the
Curie temperature.
Different phases of a material and the transitions between them were
always of greatest interest. The ancient Greek thought that everything
consists of fire, water, air and earth, which are the archetypes of
phases. Still phase transitions are alone on the basis of their ubiquity
important phenomenons. The understanding allows applications from the
refrigerator to shatter resistant mobile phone glass \cite{PJournalGlass}.
One of the simplest models with a second order
phase transition, i.e.\ a continuous transition without discontinuities of
the order parameter, is the Ising model \cite{Ising1925} in \(d \ge 2\)
dimensions. This is a simple
model of a ferromagnet and will be explained in more detail in section
\ref{ssec:isingmodel}. It is analytically solved in two dimensions on
some regular lattices \cite{Onsager1944} \cite{Wannier1945}.
In this thesis its behavior near the critical temperature -- also called
Curie temperature -- and the behavior of the critical temperature itself
on some irregular lattices corresponding to proximity graphs
(see Sec.\ \ref{ssec:graphtypes}) will be examined using the Monte
Carlo simulations described in Sec.\ \ref{ssec:montecarlo}.

>>>>

The computing capabilities of modern computers enable researchers to
collect and analyze vast amounts of data. The prime example for this fact
is the LHC, where petabytes of particle collision data are collected,
processed and discarded or kept every second \cite{LHC}.
But computers are not only able to process measured data, but
also to generate data by simulations of, e.g., statistical systems, where
the rules for each subsystem are well defined but the behavior of the
whole system is not easy to predict because of the interactions of the
subsystems. For those systems there often exists no analytic solutions
or only for very simplified or special cases. But one can simulate all
the interactions of the subsystems and observe the behavior of the whole
system.
For example one can examine phase transitions, which are defined by the abrupt
change of an observable, e.g., the change of the density of water near
boiling or the change of the magnetization of a ferromagnet near the
Curie temperature.
Different phases of a material and the transitions between them were
always of greatest interest. The ancient Greek thought that everything
consists of fire, water, air and earth, which are the archetypes of
phases. Still phase transitions are important phenomenons, because they
are ubiquitious. The understanding allows applications from the
refrigerator to shatter resistant mobile phone glass \cite{PJournalGlass}.
One of the simplest models with a second order
phase transition, i.e.\ a continuous transition without discontinuities of
the order parameter, is the Ising model \cite{Ising1925} in \(d \ge 2\)
dimensions. This is a simple
model of a ferromagnet and will be explained in more detail in Sec.\
\ref{ssec:isingmodel}. It is analytically solved in two dimensions on
some regular lattices \cite{Onsager1944} \cite{Wannier1945}.
In this thesis its behavior near the critical temperature -- also called
Curie temperature -- and the behavior of the critical temperature itself
on some irregular lattices corresponding to proximity graphs
(see Sec.\ \ref{ssec:graphtypes}) will be examined using the Monte
Carlo simulations described in Sec.\ \ref{ssec:montecarlo}.

### 2013-09-03 Tex/chapters/methods.tex
Rechtschreibfehler und Kleinigkeiten

\subsection{Thermodynamic Theory}
\label{ssec:theory}
    The disordered Ising model will be examined as a \emph{canonical system} in
    \emph{equilibrium}. A canonical system can exchange energy with a
    heat bath, thus it has a constant temperature equal to the temperature of the
    heat bath.
    Equilibrium is defined as a steady state, wherein
    the observables are only fluctuating but not changing in any
    particular direction. For example thermal equilibrium denotes the
    condition, that the system under scrutiny has reached the temperature
    of the heat bath. This is the case, once there is no directed energy
    exchange between them, thus the energy of the simulated system reaches
    a steady state, where only fluctuations occur.
    In a canonical ensemble the probability \(p_i\) of a state
    \(i\) is distributed according to a Boltzman distribution
    \begin{equation}
        p_i \propto e^{-\beta H_i}
    \end{equation}
    \begin{equation}
        \beta = \frac{1}{k_B T}
    \end{equation}
    Further the free energy \(F\) of a canonical ensemble is minimized
    in equilibrium and all observables can be derived from \(F\)
    in a straight forward way, as stated in every textbook about
    statistical physics (e.g.\ \cite{nolting2005}).\\
    Because of
    \begin{equation}
        F = U - TS
    \end{equation}
    where \(S\) is the entropy, \(U=\avg{H}\) the internal energy and
    \(\avg{\cdot}\) declares the expected value of an observable.
    One can guess that for low \(T\) the internal energy
    will be low, and for high \(T\) the entropy will be high, to minimize
    \(F\).
    Considering the Hamiltonian of the Ising model, a spin configuration
    of high order, where most spins are aligned with their neighbors,
    leads to a low \(H\) and therefore a low \(U\). Simultaneously, this
    state of high order corresponds to a low entropy \(S\).
    Analogically a state of high \(U\) is also a state of high \(S\).
    These preliminary considerations make a phase transition at some
    \(T\) where the influence of the entropy on \(F\) becomes the same
    order of magnitude as the influence of the internal energy on \(F\),
    very plausible.\\
    To determine
    \begin{equation}
        F=-k_{B}T \ln{Z}
    \end{equation}
    or any expected value of an observable
    \begin{equation}
        \avg{O} = \frac{1}{Z} \sum_i O_i e^{-\beta H_i},
    \end{equation}
    one has to know the partition function
    \begin{equation}
        Z = \sum_i p_i = \sum_i e^{-\beta H_i},
        \label{eq:partitionFunction}
    \end{equation}
    where the sum goes over all possible states \(i\) of the system.
    Because every site can have two states, there are \(2^N\) different
    states of the system. For each the energy \(H_i\) has to be calculated
    to solve the sum from Eq.\ \eqref{eq:partitionFunction} to gain \(Z\).
    Hence for small \(N\) the partition function is computable, but the system
    may show very different properties than in the thermodynamic limit.
    To minimize these finite size effects it is desirable to examine
    systems with large \(N\). But \(2^N\) is a very rapidly increasing
    number, so for large \(N\) it is unfeasible to calculate the energy for each state, except
    for cases where it is possible to solve it analytically.\\
    If not, one can get estimates of the observables for big \(N\) using
    Monte Carlo simulations, which are introduced in the next chapter.\\

    The observables which are measured in this thesis are the mean
    magnetization per spin
    \begin{equation}
        m = \frac{1}{N} \sum_i s_i
    \end{equation}
    and the mean energy per spin
    \begin{equation}
        E = \frac{1}{N} \avg{H}.
    \end{equation}
    As mentioned before, properties near the phase transition and the
    critical temperature \(T_c\), where the phase transition occurs, will
    be examined. The Ising system in two dimensions shows a second order
    phase transition, hence \(m\) and \(E\) are continous, but show at
    \(T_c\) an infinitly sharp slope in the thermodynamic limit i.e.\
    the first derivative diverges.
    From statistical physics (compare \cite{nolting2005}) is known that
    these derivatives can be expressed by fluctuations e.g.\ the specific
    heat can be expressed as
    \begin{equation}
        c = \frac{\partial \avg{H}}{\partial T} = k_B \beta^2 \avg{\brac{H-\avg{H}}^2}
    \end{equation}
    The specific heat is a measure for how much energy is needed to change
    the temperature of the system.
    Analogically the susceptibility
    \begin{equation}
        \chi = N \beta \avg{\brac{m-\avg{m}}^2}
    \end{equation}
    is a measure for how strong an outer magnetic field changes the
    magnetization of the system.
    Beside these observables from classical physics, a fifth observable
    the binder cumulant \cite{Binder1981}
    \begin{equation}
        g = \frac{3}{2}\brac{1-\frac{\avg{m^4}}{3\avg{m^2}^2}}
        \label{eq:binder}
    \end{equation}
    is considered.
    This is a dimensionless value, which can be used to determine the
    critical point. These five observables will be sufficient to analyse
    the phase transition in the scope of this bachelor thesis. All of them
    can be easily computed when \(m\) and \(E\) are measured. The next
    chapter will show, how to get estimates for them through Monte Carlo
    simulations.

\subsection{Monte Carlo Simulations}
\label{ssec:montecarlo}
    The idea behind Monte Carlo simulations is to take random samples of
    the observable, which should be measured, and to estimate the mean of
    the observable from this samples. To apply this technique to statistical
    ensembles, one creates sample states of the system, measures the
    observables calculates the expected value through averaging.\\
    In statistical physics the expected value of an observable \(O\)
    is -- as also noted above -- calculated by
    \begin{equation}
        \avg{O} = \frac{1}{Z} \sum_i p_i O_i
    \end{equation}
    It is however possible that there are few states contributing
    massively more than others. In canonical systems at low \(T\),
    states with low values of \(H\) contribute much more than states with
    high values of \(H\).
    But if one samples the \(2^N\) states evenly, it is probable to miss
    them. this is called \emph{simple sampling} and results in large
    errorbars for any observable.
    It is therefore desirable to sample only the states with high
    contributions to the sum.
    As noted before, the system under scrutiny is canonical and therefore
    \(p_i\) is known. The states are distributed according to the Boltzmann
    distribution. Hence \emph{Importance Sampling} can be utilized.\\
    Instead of sampling uniformly distributed random states, one should sample
    states according to their occurence probability given by the Boltzmann
    distribution. In fact this reduces the estimator to
    \begin{equation}
        O_M = \frac{1}{M} \sum_{i=1}^M O_i,
    \end{equation}
    where \(M\) is the number of samples. The proof is shown in \cite{NewmanBarkema1999}.
    This is a very convenient form.\\
    But it is difficult to create a random state of a physical system,
    e.g.\ the Ising system, according to a given distribution. The simple
    approach of creating uniformly distributed random states and reject
    them with probability \(p_i^{-1}\) depending on their energy is not
    efficient, because many generated states will be discarded and the
    computing time to generate them and calculating their energy will
    be wasted.
    Hence one uses \emph{Markov Chains} to generate new states \(\nu\)
    from former ones \(\mu\). It is important that the transition probabilities
    \(A(\mu \to \nu)\) obey \emph{Detailed Balance} and \emph{Ergodicity}.
    \emph{Detailed Balance} means that the probability to leave a state is
    the same as the probability to enter the state in equilibrium
    \(p_\mu A(\mu \to \nu) = p_\nu A(\nu \to \mu)\) with \(p_\mu\) the
    probability to be in state \(\mu\). This ensures that the system can
    equilibrate and that the states are distributed according to the
    desired distribution in equilibrium \cite{NewmanBarkema1999}.
    And \emph{Ergodicity} requires that every possible state is reachable
    from every other state in finite time \cite{NewmanBarkema1999} \cite{Katzgraber2011}.
    Otherwise the samples might not be representative for the whole system.\\
    In this thesis three algorithms, which fulfill all requirements,
    were used. They will be shortly described in the following subsections.

@@

        The count of sweeps till

>>>>

\subsection{Thermodynamic Theory}
\label{ssec:theory}
    The disordered Ising model will be examined as a \emph{canonical system} in
    \emph{equilibrium}. A canonical system can exchange energy with a
    heat bath, thus it has a constant temperature equal to the temperature of the
    heat bath.
    Equilibrium is defined as a steady state, wherein
    the observables are only fluctuating but not changing in any
    particular direction. For example thermal equilibrium denotes the
    condition that the system under scrutiny has reached the temperature
    of the heat bath. This is the case, once there is no directed energy
    exchange between them, thus the energy of the simulated system reaches
    a steady state, where only fluctuations occur.
    In a canonical ensemble the probability \(p_i\) of a state
    \(i\) is distributed according to a Boltzman distribution
    \begin{equation}
        p_i \propto e^{-\beta H_i}
    \end{equation}
    \begin{equation}
        \beta = \frac{1}{k_B T}
    \end{equation}
    Further the free energy \(F\) of a canonical ensemble is minimized
    in equilibrium and all observables can be derived from \(F\)
    in a straight forward way, as stated in every textbook about
    statistical physics (e.g.\ \cite{nolting2005}).\\
    Because of
    \begin{equation}
        F = U - TS
    \end{equation}
    where \(S\) is the entropy, \(U=\avg{H}\) the internal energy and
    \(\avg{\cdot}\) declares the expectation value of an observable.
    One can guess that for low \(T\) the internal energy
    will be low, and for high \(T\) the entropy will be high to minimize
    \(F\).
    Considering the Hamiltonian of the Ising model, a spin configuration
    of high order, where most spins are aligned with their neighbors,
    leads to a low value of \(H\) and therefore a low value of \(U\).
    Simultaneously, this state of high order corresponds to a low entropy \(S\).
    Analogically a state of high \(U\) is also a state of high \(S\).
    These preliminary considerations make a phase transition at some
    \(T\) where the influence of the entropy on \(F\) becomes the same
    order of magnitude as the influence of the internal energy on \(F\),
    very plausible.\\
    To determine
    \begin{equation}
        F=-k_{B}T \ln{Z}
    \end{equation}
    or any expected value of an observable
    \begin{equation}
        \avg{O} = \frac{1}{Z} \sum_i O_i e^{-\beta H_i},
    \end{equation}
    one has to know the partition function
    \begin{equation}
        Z = \sum_i p_i = \sum_i e^{-\beta H_i},
        \label{eq:partitionFunction}
    \end{equation}
    where the sum goes over all possible states \(i\) of the system.
    Because every site can have two states, there are \(2^N\) different
    states of the system. For each the energy \(H_i\) has to be calculated
    to solve the sum from Eq.\ \eqref{eq:partitionFunction} to gain \(Z\).
    Hence for small \(N\) the partition function is computable, but the system
    may show very different properties than in the thermodynamic limit.
    To minimize these finite size effects it is desirable to examine
    systems with large \(N\). But \(2^N\) is a very rapidly increasing
    number, so for large \(N\) it is unfeasible to calculate the energy for each state, except
    for cases where it is possible to solve it analytically.\\
    If not, one can get estimates of the observables for big \(N\) using
    Monte Carlo simulations, which are introduced in the next chapter.\\

    The observables which are measured in this thesis are the mean
    magnetization per spin
    \begin{equation}
        m = \frac{1}{N} \sum_i s_i
    \end{equation}
    and the mean energy per spin
    \begin{equation}
        E = \frac{1}{N} \avg{H}.
    \end{equation}
    As mentioned before, properties near the phase transition and the
    critical temperature \(T_c\), where the phase transition occurs, will
    be examined. The Ising system in two dimensions shows a second order
    phase transition, hence \(m\) and \(E\) are continous, but show at
    \(T_c\) an infinitly sharp slope in the thermodynamic limit, i.e.\
    the first derivative diverges.
    From statistical physics (compare \cite{nolting2005}) it is known that
    these derivatives can be expressed by fluctuations, e.g.\ the specific
    heat can be expressed as
    \begin{equation}
        c = \frac{\partial \avg{H}}{\partial T} = k_B \beta^2 \avg{\brac{H-\avg{H}}^2}.
    \end{equation}
    The specific heat is a measure for how much energy is needed to change
    the temperature of the system.
    Analogically the susceptibility
    \begin{equation}
        \chi = N \beta \avg{\brac{m-\avg{m}}^2}
    \end{equation}
    is a measure for how strong an outer magnetic field changes the
    magnetization of the system.
    Beside these observables from classical physics, a fifth observable
    the binder cumulant \cite{Binder1981}
    \begin{equation}
        g = \frac{3}{2}\brac{1-\frac{\avg{m^4}}{3\avg{m^2}^2}}
        \label{eq:binder}
    \end{equation}
    is considered.
    This is a dimensionless value, which can be used to determine the
    critical point. These five observables will be sufficient to analyse
    the phase transition in the scope of this bachelor thesis. All of them
    can be easily computed when \(m\) and \(E\) are measured. The next
    chapter will show, how to get estimates for them through Monte Carlo
    simulations.

\subsection{Monte Carlo Simulations}
\label{ssec:montecarlo}
    The idea behind Monte Carlo simulations is to take random samples of
    the observable, which should be measured, and to estimate the mean of
    the observable from this samples. To apply this technique to statistical
    ensembles, one creates sample states of the system, measures the
    observables and calculates the expected value through averaging.\\
    In statistical physics the expected value of an observable \(O\)
    is -- as also noted above -- calculated by
    \begin{equation}
        \avg{O} = \frac{1}{Z} \sum_i p_i O_i.
    \end{equation}
    It is however possible that there are few states contributing
    massively more than others. In canonical systems at low \(T\),
    states with low values of \(H\) contribute much more than states with
    high values of \(H\).
    But if one samples the \(2^N\) states evenly, it is probable to miss
    them. This is called \emph{simple sampling} and results in large
    errorbars for any observable.
    It is therefore desirable to sample only the states with high
    contributions to the sum.
    As noted before, the system under scrutiny is canonical and therefore
    \(p_i\) is known. The states are distributed according to the Boltzmann
    distribution. Hence \emph{Importance Sampling} can be utilized.\\
    Instead of sampling uniformly distributed random states, one should sample
    states according to their occurence probability given by the Boltzmann
    distribution. In fact this reduces the estimator to
    \begin{equation}
        O_M = \frac{1}{M} \sum_{i=1}^M O_i,
    \end{equation}
    where \(M\) is the number of samples. The proof is shown in \cite{NewmanBarkema1999}.
    This is a very convenient form.\\
    But it is difficult to create a random state of a physical system,
    e.g.\ the Ising system, according to a given distribution. The simple
    approach of creating uniformly distributed random states and reject
    them with probability \(p_i^{-1}\) depending on their energy is not
    efficient, because many generated states will be discarded and the
    computing time to generate them and calculating their energy will
    be wasted.
    Hence one uses \emph{Markov Chains} to generate new states \(\nu\)
    from former ones \(\mu\). It is important that the transition probabilities
    \(A(\mu \to \nu)\) obey \emph{Detailed Balance} and \emph{Ergodicity}.
    \emph{Detailed Balance} means that the probability to leave a state is
    the same as the probability to enter the state in equilibrium
    \(p_\mu A(\mu \to \nu) = p_\nu A(\nu \to \mu)\) with \(p_\mu\) the
    probability to be in state \(\mu\). This ensures that the system can
    equilibrate and that the states are distributed according to the
    desired distribution in equilibrium \cite{NewmanBarkema1999}.
    And \emph{Ergodicity} requires that every possible state is reachable
    from every other state in finite time \cite{NewmanBarkema1999} \cite{Katzgraber2011}.
    Otherwise the samples might not be representative for the whole system.\\
    In this thesis three algorithms, which fulfill all requirements,
    were used. They will be shortly described in the following subsections.

@@

        The count of sweeps until

### 2013-09-03 Tex/chapters/model.tex
Rechtschreibfehler und Kleinigkeiten

    For the square lattice Ising model \(J_{ij}=1 \forall i,j\) applies.

@@

    The most important modification of the square lattice Ising modell in this
    thesis is, that the sites of the square lattice are displaced.
    The displacement is randomly Gaussian distributed with the standard

@@

    displacement. If only the edges between these neighbors remain,

@@

    leaves the edges as they were before the displacement, the edges would

@@

    would not be compareable with known results of the two dimensional

@@

    square lattice Ising model. For this exists an analytic solution \cite{Onsager1944}.

@@

    and the critical exponents \(\alpha = 0\), \(\beta = \frac{1}{8}\)

@@

\subsection{Gabriel- and Relative Neighborhood Graph}
\label{ssec:graphtypes}
    A graph \(G(V,E)\) is a set of nodes \(V\) and edges \(E\). In the
    scope of this model, the nodes get identified with the sites of the
    lattice and the edges signal the neighbor status of two sites. If the
    corresponding nodes are connected, the sites are neighbors.\\
    All here mentioned graph types are \emph{proximity graphs}.
    The edges of these graphs connect nodes which are by some metric near
    to each other.
    Hence they are suited to generalize problems defined on regular
    lattices with nearest neighbor relationships.
    In this thesis the distance is always determined by the Euclidean
    metric in two dimensions, though in principle every metric in any
    dimension can be used.\\

@@

    \(d_{ij}\) are connected with an edge, if a circle with its

@@

    distance \(d_{ij}\) are connected, if no other node is in the
    \emph{lune}. The lune is defined as the interSec.\ of two

@@

    To construct these graphs the most simple way is to test for each

>>>>

    For the square lattice Ising model \(J_{ij} = 1 \ \forall i,j\) applies.

@@

    The most important modification of the square lattice Ising model in this
    thesis is that the sites of the square lattice are displaced.
    The displacement is randomly Gaussian distributed with standard

@@

    displacement. If only the edges between these neighbors remained,

@@

    left the edges as the were before the displacement, the edges would

@@

    would not be comparable with known results for the two dimensional

@@

    square lattice Ising model. For this an analytic solution exists \cite{Onsager1944}.

@@

    and the critical exponents are \(\alpha = 0\), \(\beta = \frac{1}{8}\)

@@

\subsection{Gabriel- and Relative Neighborhood Graph}
\label{ssec:graphtypes}
    A graph \(G(V,E)\) is a set of nodes \(V\) and edges \(E\). In the
    scope of this model, the nodes get identified with the sites of the
    lattice and the edges signal the neighbor status of two sites. If the
    corresponding nodes are connected, the sites will be neighbors.\\
    All here mentioned graph types are \emph{proximity graphs}.
    The edges of these graphs connect nodes which are by some metric near
    to each other.
    Hence they are suited to generalize problems defined on regular
    lattices with nearest neighbor relationships.
    In this thesis the distance is always determined by the Euclidean
    metric in two dimensions, though in principle every metric in any
    dimension can be used.\\

@@

    \(d_{ij}\) will be connected with an edge, if a circle with its

@@

    distance \(d_{ij}\) will be connected, if no other node is in the
    \emph{lune}. The lune is defined as the intersection of two

@@

    The simplest way to construct these graphs is to test for each
